DOpElib: Differential Equations\\ and Optimization Environment

### 2013-11-15 Examples/Documentation/introduction.tex
contents tex in opt problems

\subsection{Problem description}
In order to allow our algorithms the automatic assembly of all required 
data we need to have some container which contains the complete problem 
description in a common data format. For this we have the following 
classes in \texttt{DOpEsrc/container}
\begin{itemize}
  \item \texttt{pdeproblemcontainer.h} Is used to describe  stationary PDE problems.
  \item \texttt{instatpdeproblemcontainer.h} {\bf TODO is still missing but will be used for nonstationary problems.} This will be implemented once we have nonstationary optimization problems running to avoid error duplication in the coding process.
  \item \texttt{optproblemcontainer.h} Is used to describe  OPT problems governed by 
    stationary PDEs. 
  \item \texttt{instatoptproblemcontainer.h} Is used to describe  OPT problems governed by nonstationary PDEs. The only difference to the stationary case is that we need to specify a time-stepping method.  
\end{itemize}
In order to fill these containers there are two things to be done,
first we need to actually write some data, for instance,
the semilinear form $a(\cdot)(\cdot)$, a target functional $J(\cdot)$, etc.,
which describe the problem. Then we have to select some numerical 
algorithm components like finite elements, linear solvers $\ldots$.
The latter ones should be written such that when exchanging these components
none of the problem descriptions should require changes. 
Note that it still may be necessary to write some additional descriptions, 
e.g., if you solve the PDE with a fix point iteration you don't need derivatives
but if you want to use Newton's method, derivatives are needed.

@@

\subsubsection{Space-time handler}
First we need to select a method how to handle all dofs in space and time.
\begin{itemize}
\item \texttt{basic/spacetimehandler\underline{ }base.h} This class is used to define 
  an interface to the dimension independent functionality of all space time dof handlers.
  {\bf TODO: Beispiele geben}
\item \texttt{basic/statespacetimehandler.h} Another intermediate interface class which adds 
  the dimension dependent functionality if only the variable $u$ is considered, i.e., a 
  PDE problem.
\item \texttt{basic/spacetimehandler.h } Same as above but with both $q$ and $u$, i.e., for
  OPT problems.
\item \texttt{basic/mol\underline{ }statespacetimehandler.h} Implementation of a method of 
  line space time dof handler for PDE problems. It has only one spatial 
  dofhandler that is used for all time intervals.
\item \texttt{basic/mol\underline{ }spacetimehandler.h} Same as above for OPT problems.
  A separate spatial dof handler for each of the variables $q$ and $u$ is maintained 
  but only one triangulation.
\item \texttt{basic/mol\underline{ }multimesh\underline{ }spacetimehandler.h}
  Same as above, but now in addition the triangulations for $q$ and $u$ can be refined
  separately from one common initial coarse triangulation. Note that this will
  in addition require the use of the multimesh version for integrator and 
  face- as well as celldatacontainer.
\end{itemize}
Note that we use these for stationary problems as well, but then you don't have to specify
any time information.

>>>>

\subsection{Problem description}
In order to allow our algorithms the automatic assembly of all required 
data we need to have some container which contains the complete problem 
description in a common data format. For this we have the following 
classes in \texttt{DOpEsrc/container}
\begin{itemize}
  \item \texttt{pdeproblemcontainer.h} Is used to describe  stationary PDE problems.
  \item \texttt{instatpdeproblemcontainer.h} 
%{\bf TODO is still missing but will be used for nonstationary problems.} 
This will be implemented once we have nonstationary optimization problems running to avoid error duplication in the coding process.
  \item \texttt{optproblemcontainer.h} Is used to describe  OPT problems governed by 
    stationary PDEs. 
  \item \texttt{instatoptproblemcontainer.h} Is used to describe  OPT problems governed by nonstationary PDEs. The only difference to the stationary case is that we need to specify a time-stepping method.  
\end{itemize}
In order to fill these containers there are two things to be done,
first we need to actually write some data, for instance,
the semilinear form $a(\cdot)(\cdot)$, a target functional $J(\cdot)$, etc.,
which describe the problem. Then we have to select some numerical 
algorithm components like finite elements, linear solvers $\ldots$.
The latter ones should be written such that when exchanging these components
none of the problem descriptions should require changes. 
Note that it still may be necessary to write some additional descriptions, 
e.g., if you solve the PDE with a fix point iteration you don't need derivatives
but if you want to use Newton's method, derivatives are needed.

@@

\subsubsection{Space-time handler}
First we need to select a method how to handle all dofs in space and time.
\begin{itemize}
\item \texttt{basic/spacetimehandler\underline{ }base.h} This class is used to define 
  an interface to the dimension independent functionality of all space time dof handlers.
  %{\bf TODO: Beispiele geben}
\item \texttt{basic/statespacetimehandler.h} Another intermediate interface class which adds 
  the dimension dependent functionality if only the variable $u$ is considered, i.e., a 
  PDE problem.
\item \texttt{basic/spacetimehandler.h } Same as above but with both $q$ and $u$, i.e., for
  OPT problems.
\item \texttt{basic/mol\underline{ }statespacetimehandler.h} Implementation of a method of 
  line space time dof handler for PDE problems. It has only one spatial 
  dofhandler that is used for all time intervals.
\item \texttt{basic/mol\underline{ }spacetimehandler.h} Same as above for OPT problems.
  A separate spatial dof handler for each of the variables $q$ and $u$ is maintained 
  but only one triangulation.
\item \texttt{basic/mol\underline{ }multimesh\underline{ }spacetimehandler.h}
  Same as above, but now in addition the triangulations for $q$ and $u$ can be refined
  separately from one common initial coarse triangulation. Note that this will
  in addition require the use of the multimesh version for integrator and 
  face- as well as celldatacontainer.
\end{itemize}
Note that we use these for stationary problems as well, but then you don't have to specify
any time information.

### 2013-11-15 Examples/OPT/StatPDE/Example3/content.tex
Changed DOpE to DOpElib in the documentation, finished overhaul of PDEStatPDE-Examples
Example 3 laeuft
contents tex in opt problems

In this example, we solve a optrimization problems from 
fluid dynamcis. The configuration is similar to 
the fluid optimization problem proposed by Roland Becker 
in 2000 ``Mesh adaption for stationary flow control''.

We describe the underlying equations later...

>>>>

\subsubsection{General problem description}
In this example, we solve a optimization problems from 
fluid dynamcis. The configuration is similar to 
the fluid optimization problem proposed by Roland Becker 
``Mesh adaption for stationary flow control'' (2000).

The configuration
    comes from the original fluid benchmark problem
    and has been modified to reduce drag around the
    cylinder. To gain the solvability of
    the optimization problem we add a quadratic
    regularization term to the cost functional.

\begin{figure}[h]
\centering
    {\includegraphics[width=7cm]{Documentation/visit_Nov_15_2013_0000_a.png}}
    {\includegraphics[width=7cm]{Documentation/visit_Nov_15_2013_0001_a.png}}
  \caption{Configuration of the cylinder-drag minimization problem. 
By sucking out the fluid (right the $y$-velocity), 
the force on the cylinder is reduced. At left, the $x$-velocity field 
is shown. Behind the cylinder, almost no fluid goes from left to right, 
which is shown in blue color.}
  \label{ex_2_plots}
\end{figure}

The computational domain is denoted by $\Omega$ with the 
boundary $\partial\Omega
= \Gamma_in \cup \Gamma_out \cup \Gamma_Q \cup \Gamma_w$, 
where $\Gamma_in$ and $\Gamma_out$ denote the inflow and outflow
boundaries, respectively. On $\Gamma_in$, we prescribe a fixed parabolic 
inflow profile. The part(s) $\Gamma_w$ denote the top and bottom 
boundaries. Finally, $\Gamma_Q$, represent Neumann control boundaries. Here,
we prescribe 
\[
\rho\nu \partial_n v - pn = qn \quad\text{on } \Gamma_Q
\] 
where $n$ denotes the outer unit normal to $\Gamma_Q$.

The state equations are given by: Find $v$ and $p$ such that
\begin{align*}
- \nabla\cdot \sigma(v,p) + v\cdot \nabla v = 0,\\
\nabla\cdot v = 0
\end{align*}
with $\sigma(v,p) = -pI + \rho\nu(\nabla v + \nabla v^T)$.

\textbf{Remark:} Since, we use the 
symmetric stress tensor, we need to subtract the nonsymmetric part 
on the outflow boundary, related to the do-nothing condition.
$\diamond$

The target functional is considered as 
\[
k(v,p) = \int_{\Gamma_O} n\cdot \sigma(v,p)\cdot d \, ds,
\]
where $\Gamma_O$ denotes the cylinder boundary, and $d$ is a vector in the
direction
of the mean flow. For theoretical and numerical reasons, this functional 
needs to be regularized, including the control variable $q$, such that
\[
K(v,p,q) = k(v,p) + \frac{\alpha}{2}||q - q_0||,
\] 
where $\alpha$ is the Tikhonov parameter and $q_0$ some 
reference control.

The rest of the progam is similar to the previous optimization problems where
we formulate the state equation in a weak form $a(v,p)(\phi)$ such that the 
final problem reads
\[
K(v,p,q) \rightarrow 0 \quad \text{s.t.} \quad a(v,p)(\phi) = 0.
\]

%\subsubsection{Program description}

### 2013-11-15 Examples/OPT/StatPDE/Example9/content.tex
contents tex in opt problems

\subsubsection{General program description}
In this example we solve an optimization problem where 
the equations come from fluid-structure interaction (FSI). The idea
is to extend the steady FSI benchmark problem (FSI 1, proposed by Hron/Turek) to an optimization 
problem where the drag is minimized over the cylinder and the beam.

The example has many similarities to Opt Example 3.

Detailed description will follow later. %TODO

>>>>

\subsubsection{General program description}
In this example we solve an optimization problem where 
the equations come from fluid-structure interaction (FSI). The idea
is to extend the steady FSI benchmark problem (FSI 1, proposed by Hron/Turek) 
to an optimization 
problem where the drag is minimized over the cylinder and the beam. 
The setting is similar to Opt Example 3. In fact, the only 
novel things are to attach an elastic beam at the cylinder and to 
extend the equation to fluid-structure interaction (instead of 
pure fluid as in Example 3).

\begin{figure}[h]
\centering
    {\includegraphics[width=7cm]{Documentation/visit_fsi_1_opt_control_p3_0002_bearbeitet_Aug_15.png}}
    {\includegraphics[width=7cm]{Documentation/visit_fsi_1_opt_control_p3_0005_bearbeitet_Aug_15.png}}
  \caption{Configuration of the fsi cylinder-beam-drag minimization problem. 
By sucking out the fluid, 
the force on the cylinder is reduced. At left, the $x$-velocity field 
is shown. In right figure, we the corresponding adjoint solution is shown.}
  \label{ex_2_plots}
\end{figure}

The state equation system reads:
\begin{Problem}[Stationary Fluid-Structure Interaction with STVK material]
Let $q$ denote the control variable. Find $U:= \{\hat v,\hat p,\hat u\}$ such that
  \begin{equation*}
    \begin{aligned}
      (\hat J \rho_f \hat F^{-1}\hat  v\cdot\hat\nabla \hat v,
      \hat\phi^v)_{\hat\Omega_f}
      + (\hat J\hat\sigma_f\hat F^{-T},\hat \nabla\hat\phi^v)_{\hat\Omega_f}\\
      + (\hat J\hat\sigma_s\hat F^{-T},\hat \nabla\hat\phi^v)_{\hat\Omega_s}
      + \langle q,\hat\phi^v\cdot \hat n\rangle_{\Gamma_Q}
      &= 0&&\forall\hat\phi^v\in \hat V,\\
      %%%%%%%%%%%% 
      (\hat v,\hat\phi^u)_{\hat\Omega_s}
      + (\alpha_u \hat \nabla \hat u,\hat \nabla\hat\phi^u)_{\hat\Omega_f}
      &=0&&\forall\hat\phi^u\in \hat V,\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      (\widehat{\text{div}}\,(\hat J\hat F^{-1}\hat
      v_f),\hat\phi^p)_{\hat\Omega_f} 
      + (\hat p,\hat \phi^p)_{\hat\Omega_s}
      &=0&&\forall\hat \phi^p\in \hat L,
    \end{aligned}
  \end{equation*}  
\end{Problem}

The target functional is considered as 
\[
k(U) = \int_{\hat\Gamma_O \cup \hat\Gamma_i} \hat n\cdot \hat J\hat\sigma(v,p)\hat
F^{-T} \cdot \hat d \, ds,
\]
where $\Gamma_O$ denotes the cylinder boundary and $\Gamma_i$ the 
interface between fluid and solid, and $\hat d$ is a vector in the
direction
of the mean flow. Moreover, $\hat J$ and $\hat F$ denote the deformation 
gradient and its determinant as well known in fluid-structure interaction.
For theoretical and numerical reasons, this functional 
needs to be regularized, including the control variable $q$, such that
\[
K(U,q) = k(U) + \frac{\alpha}{2}||q - q_0||,
\] 
where $\alpha$ is the Tikhonov parameter and $q_0$ some 
reference control.

The rest of the progam is similar to the previous optimization problems where
we formulate the state equation in a weak form $a(v,p)(\phi)$ such that the 
final problem reads
\[
K(U,q) \rightarrow 0 \quad \text{s.t.} \quad a(v,p)(\phi) = 0.
\]
